Getting Started with the Graded Response Model (GRM): An introduction and tutorial in R

### 2024-08-19 manuscript.qmd
Revision round 1

---
title: "Getting Started with the Graded Response Model (GRM): A gentle introduction and tutorial in R"
shorttitle: "The Graded Response Model Tutorial"
author:
  - name: Rizqy Amelia Zein
    corresponding: true
    orcid: 0000-0001-7840-0299
    email: amelia.zein@psikologi.unair.ac.id
    affiliations:
      - name: Universitas Airlangga
        department: Department of Psychology
        address: Jalan Airlangga 4-6
        city: Surabaya
        region: Jawa Timur, Indonesia
        postal-code: "60286"
      - name: Ludwig-Maximilians-Universität
        department: Department of Psychology
        address: Leopoldstraße 13
        city: Munich
        postal-code: "80802"
  - name: Hanif Akhtar
    orcid: 0000-0002-1388-7347
    affiliations:
      - name: ELTE Eötvös Loránd University
        department: Doctoral School of Psychology
        country: Hungary
      - name: Universitas Muhammadiyah Malang
        department: Faculty of Psychology
        city: Malang
        region: Jawa Timur, Indonesia
abstract: "This tutorial introduces the Graded Response Model (GRM), a tool for testing measurement precision within the Item Response Theory (IRT) paradigm. Addressing common problems of measurement imprecision and lack of construct validity, the tutorial guides researchers through a unidimensional GRM analysis in the R environment, using psych, mirt, and ggmirt packages. GRM is specifically designed to examine the psychometric properties of psychological scales with polytomous (Likert-style) items. The tutorial illustrates the procedure using data from the Open Psychometrics Database on the Right-Wing Authoritarianism (RWA) scale, outlining the theoretical underpinnings of GRM, and steps for data preparation, testing model assumptions, model fitting, and interpretation of results."
keywords:
  - graded response model
  - item response theory
  - mirt
  - ggmirt
  - psych
  - R
author-note:
  disclosures:
    conflict-of-interest: "We declare we have no competing interests."
    financial-support: "The first author receives a PhD scholarship from the German Academic Exchange Service (Deutscher Akademischer Austauschdienst, DAAD)."
lang: en
bibliography: references.bib
format:
  apaquarto-docx: default
editor: visual
execute: 
  warning: false
  error: false
  output: false
  echo: true
---

@@

install.packages("flextable", dependencies=TRUE) 
library(flextable)
```

Only half of published findings include evidence of measurement validity [@flakeConstructValidationSocial2017] and a large proportion of published studies suffer from hidden invalidity [@husseyHiddenInvalidity152020]. Lack of evidence for measurement validity can raise serious concerns about the replicability of research findings, as it directly contributes to the invalidity of the findings [@flakeMeasurementSchmeasurementQuestionable2020; @lilienfeldPsychologicalMeasurementReplication2020]. Furthermore, the lack of evidence for measurement validity can hamper consensus on the best way to measure a construct, which becomes a major obstacle to the establishment of cumulative psychological science [@elsonPsychologicalMeasuresAren2023]. Given these circumstances, it is critical to provide applied psychological researchers with a set of tools and practical steps to help them assess measurement validity so that they can ensure the credibility of their findings.

The purpose of this article is to provide a brief introduction to the Graded Response Model (GRM), a family of item response models specifically designed to assess the measurement precision of a polytomous (Likert-style) scale. We begin this paper by briefly explaining the basics of Item Response Theory (IRT) and contrasting its differences with the widely known Classical True Score Theory (CTT). In this paper, we aim to keep our tutorial concise, so we do not cover all IRT concepts in detail. Interested readers are referred to available didactic texts on IRT, such as @embretsonItemResponseTheory2000 for the introductory level, and @bakerBasicsItemResponse2017 and @deayalaTheoryPracticeItem2022 for a more technical and comprehensive overview.

Further, we briefly explain what GRM is and how it is specified. In this part, we highlight how model and item parameters are specified and interpreted as well as visualizations derived from a GRM analysis. We also introduce three key assumptions underlying a GRM and how to test these assumptions. We then provide an illustrative example of testing measurement precision using a GRM analysis, and in this part, we show how to practically implement a GRM analysis and interpret its results. We complement the illustrative example with code in the open-source programming language R [@rcoreteamLanguageEnvironmentStatistical2023] so that the readers can implement and reproduce the example presented in this paper. Readers with some experience in R and RStudio can easily follow this tutorial, but for those who are not yet familiar with R, we refer them to excellent, easy-to-follow materials developed by @navarroPsychologicalScience2018. Finally, we present some concluding remarks to encourage researchers to implement a GRM analysis as a part of their routine.

@@

To scrutinize their measurement quality, psychological researchers have relied on two basic principles underlying psychological testing, i.e., CTT and IRT, to decompose observed scores into their *deterministic* and *random* or stochastic components [@zumboValidityFoundationalIssues2006]. CTT, on the one hand, decomposes observed outcomes into *true* scores and *error* components. In practice, researchers implement this principle by performing structural equation modeling (SEM), exploratory factor analysis (EFA), or confirmatory factor analysis (CFA). Specifying the measurement part of a SEM model, or estimating EFA and CFA models, allows the covariance matrix of a given set of data to be decomposed into latent factors, which represent the true score or trait level, and error terms (i.e., residuals or unique variances). Consequently, the relationship between observed score and true score is *linear*, since true score is essentially a linear transformation of observed score [@embretsonItemResponseTheory2000]. IRT, on the other hand, models this relationship in a *probabilistic* rather than a linear fashion. More specifically, an IRT model assumes a probabilistic relationship between the observed score and the latent trait being measured ($\theta$), and in particular, this probability takes into account item properties such as item discrimination, difficulty, and, when appropriate, guessing [@bakerBasicsItemResponse2017].

The most critical distinction between CTT and IRT is how they conceptualize measurement error. In CTT, standard deviation of the observed score (*s*) is used to calculate standard error of measurement. Therefore, CTT assumes that measurement error is sample-dependent and constant for all individuals in the sample regardless of their trait level. This shortcoming limits the utility of CTT, especially since psychologists often need to interpret individual scores, rather than just evaluate the test as a whole. Additionally, since standard error of measurement is also a function of reliability, reliability is assumed to be the same across different levels of a trait or ability. This assumption is unrealistic since researchers often encounter situations where a test is overly difficult for a group of low-performing participants, providing little information beyond an indication that their trait or ability level is significantly lower than what the test can measure.

IRT offers an elegant solution to this issue by allowing the calculation of standard errors for each individual (*SE*$\theta$), and thus, test reliability can be inferred from the average of *SE*$\theta$ across a sample of participants [@embretsonItemResponseTheory2000; @langSciencePracticeItem2021]. This approach allows for the estimation of reliability at varying levels of trait or ability (i.e., test information function - TIF). By doing this, when a test is overly challenging for low-performing individuals or too easy for high-performing individuals, IRT can help identify the levels of trait or ability at which the test is most reliable. A concrete example of this is a study that examined the reliability of the Short Dark Tetrad (SD4) scale from an IRT perspective [@blotnerDarkTriadDead2022]. According to this study, the sadism sub-scale of the SD4 is the most reliable for measuring individuals with average to high levels ($\bar{x}$ \< $\theta$ \< 2.5 *SD*) of sadism but suboptimal for measuring individuals with lower levels of sadism [@blotnerDarkTriadDead2022]. In this sense, IRT models offer a more informative and nuanced perspective that proves useful for researchers who wish to closely examine the performance of their measures.

Further, IRT allows researchers to examine the performance of a specific item by specifying the relationship between item score and $\theta$ given one-, two-, or three-parameter. The one-parameter logistic (1PL) or a Rasch model, the simplest variant of the IRT models, accounts for only item difficulty (*b*) while assuming equal item discrimination (*a*) across items. In practice, this model overlaps slightly with CTT in the sense that they assume that all items have equal informative value, so that the estimated $\theta$ of a 1PL/Rasch model is identical to the sum score of CTT [@langSciencePracticeItem2021; @stemlerRaschMeasurementItem2021]. This notion actually stems from a philosophical principle of *specific objectivity*, which requires that comparisons between measures remain independent of both the item and individual characteristics [@raschSpecificObjectivityAttempt1977]. Put simply, a Rasch model assumes that the latent trait ($\theta$) should remain unaffected by specific items used in the test. More concretely, the difference in depression levels between two individuals should always be the same regardless of the scale used to measure their depression levels. While this principle reasonably enforces objectivity in measurement practices, it requires a strong theory underlying the construct and strict requirements of data-model fit, and both of these assumptions rarely hold in a real-world scenario, especially in psychological science.

Furthermore, the two-parameter logistic (2PL) model fills this gap by allowing for differences in item ability to differentiate individuals with varying levels of $\theta$ (i.e., item discrimination parameter - *a*). In addition, in some contexts, researchers may suspect that some parts of the probabilistic relationship between observed score and $\theta$ is explained by guessing, thus three-parameter logistic (3PL) model incorporates guessing parameter (*c*). The field of IRT has rapidly evolved to include various models suited to specific contexts, such as handling ordinal responses [@murakiGeneralizedPartialCredit1992; @samejimaGradedResponseModel1997], categorical responses [@thissenNominalCategoriesItem2013], or assessing multidimensional traits simultaneously [@bockMarginalMaximumLikelihood1981; @chalmersMirtMultidimensionalItem2012]. It is important to note that 1PL, 2PL, and 3PL models are only applicable to binary or dichotomous data (e.g., true/false responses), and the focus of this article is nonetheless to demonstrate the utility of IRT for fitting the ordered (Likert-style) responses. We briefly summarize the features of the most commonly used IRT models in {apatb-table1}.

@@

```{r}
#| label: apatb-table1
#| apa-cap: Comparison Between The Commonly Used IRT Models
#| apa-note: NULL
#| echo: false
```

+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+-------------------------+
| Model                                   | Key Characteristics                                                                                                                                         | Data Type   | Response Options        |
+=========================================+=============================================================================================================================================================+=============+=========================+
| 1-PL Model (Rasch Model)                | 1\. Estimates only item difficulties (*b*).                                                                                                                 | Dichotomous | Correct/Incorrect (0/1) |
|                                         |                                                                                                                                                             |             |                         |
|                                         | 2\. Assumes that all items have equal discrimination parameters (*a*).                                                                                      |             |                         |
|                                         |                                                                                                                                                             |             |                         |
|                                         | 3\. Item and person parameters are independent.                                                                                                             |             |                         |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+-------------------------+
| 2-PL Model                              | 1\. Estimates item difficulties (*b*) and item discrimination (*a*).                                                                                        | Dichotomous | Correct/Incorrect (0/1) |
|                                         |                                                                                                                                                             |             |                         |
|                                         | 2\. Less stringent than 1-PL model since it allows item discrimination parameters (*a*) to vary.                                                            |             |                         |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+-------------------------+
| 3-PL Model                              | 1\. Estimates item difficulties (*b*), discrimination (*a*), and pseudo-guessing parameter (*c*).                                                           | Dichotomous | Correct/Incorrect (0/1) |
|                                         |                                                                                                                                                             |             |                         |
|                                         | 2\. Appropriate for modeling a test data with multiple responses (e.g., multiple-choice tests), and thus, guessing might influence participants' responses. |             |                         |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+-------------------------+
| Graded Response Model (GRM)             | 1\. Appropriate for modeling ordinal data with more than two response categories (i.e., Likert-style).                                                      | Polytomous  | Ordered Categories      |
|                                         |                                                                                                                                                             |             |                         |
|                                         | 2\. Estimates a discrimination parameter (*a*) and multiple threshold parameters (*b*) per item.                                                            |             |                         |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+-------------------------+
| Partial Credit Model (PCM)              | 1\. An extension of the 1-PL (Rasch) model for polytomous items.                                                                                            | Polytomous  | Ordered Categories      |
|                                         |                                                                                                                                                             |             |                         |
|                                         | 2\. Estimates thresholds between adjacent categories but assumes equal discrimination across items.                                                         |             |                         |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+-------------------------+
| Generalized Partial Credit Model (GPCM) | Extending PCM to allow differential discrimination parameters across items.                                                                                 | Polytomous  | Ordered Categories      |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+-------------------------+
| Nominal Response Model (NRM)            | 1\. Appropriate for modeling categorical responses with no order.                                                                                           | Categorical | Unordered Categories    |
|                                         |                                                                                                                                                             |             |                         |
|                                         | 2\. Estimates discrimination parameters (*a*) and multiple category-specific parameters (*b*).                                                              |             |                         |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+-------------------------+

@@

The GRM is a family of IRT models specifically designed to analyze ordered polytomous (Likert-style) data [@samejimaGradedResponseModel1997; @samejimaGeneralGradedResponse2010; @samejimaGradedResponseModels2016]. The fundamental idea of GRM is to extend the logic of simpler dichotomous IRT models to polytomous scales by applying a probabilistic function to each response category. In dichotomous 2PL IRT models, the probability of answering a correct response is modeled as a function of $\theta$, item difficulty (*b*), and item discrimination (*a*). GRM extends this idea by modeling the probability of selecting certain response category or higher on an item (i.e., *step function*) given to item difficulty (*b*) and discrimination (*a*). For each response category (e.g., answering "agree" on a five-point Likert scale), GRM calculates the cumulative probability of a participant answering "agree" or above, given to their $\theta$. Therefore, the notion of item difficulty in dichotomous IRT models is extended to the step function (i.e., item threshold - *b*) to handle ordered polytomous data.

To illustrate the step function, consider a scale measuring sadistic personality (e.g., "*watching a fist-fight excites me*," etc.) with response categories ranging from 1 (*strongly disagree*) to 4 (*strongly agree*). Imagine these categories are hierarchically ordered from the lowest category (*strongly disagree*) to the highest (*strongly agree*), like a staircase. GRM aims to estimate the level of sadistic personality ($\theta$) required for "stepping" from one response category to another (e.g., *moderately agree* to *strongly agree*). Therefore, a GRM model of a five-point Likert scale calculates four item threshold parameters (*b*): the location of $\theta$ level where individuals are equally likely to respond 1 or 2 (*b*~1~), 2 or 3 (*b*~2~), 3 or 4 (*b*~3~), and 4 or 5 (*b*~4~). Each threshold is exactly the point where a participant is equally likely to respond to either of the adjacent response categories. For example, if *b*~1~ = -1.23 for the item "*watching a fist-fight excites me*", it implies that participants with sadism level 1.23 below the mean are equally likely to either answer *strongly disagree* or *moderately disagree*.

{apafg-fig1} shows an example of an item probability function (IPF) from an item on a scale with five response categories. In general, IPF describes a relationship between $\theta$ (x-axis) and the probability of endorsing a response category (y-axis - *P*$\theta$). {apafg-fig1} consists of five category probability curves, each of which represents the probability of endorsing a response category given $\theta$. Item threshold is exactly the location of $\theta$ where two adjacent category probability curves cross each other.

@@

```{r}
#| label: apafg-fig1
#| apa-cap: An Item Probability Function from a GRM Model
#| apa-note: NULL
#| output: true
#| echo: false

@@

The second key assumption pertains to the unidimensionality of $\theta$. This implies that the item pool tested using a GRM analysis should represent only one latent trait ($\theta$). Although it is indeed possible to account for multiple $\theta$ in an IRT model (multidimensional IRT, see @bockFullInformationItemFactor1988, we limit the scope of our tutorial to the unidimensional GRM. A unidimensional IRT model can still be robustly applied to multidimensional data if multiple latent traits are moderately intercorrelated or when there is a strong general factor (*g*) underlying the data [@reiseEvaluatingImpactMultidimensionality2015]. To test whether an item pool represents only one $\theta$, researchers can apply a parallel analysis [@guoAssessingDimensionalityIRT2023] or a factor analysis [@hambletonAssessingDimensionalitySet1986], before specifying their GRM models.

Relatedly, a GRM model assumes that the item pool is locally independent, as the third assumption. Local independence suggests that participants' responses to one item do not influence their response to another. It is important to note that evidence for unidimensionality does not always warrant local independence. As an eyeball example - consider a scale measuring trust in science, which contains: "*I trust scientists working in natural science*" and "*Most biologists are trustworthy*." While both items might measure the same latent trait (i.e., trust in science), those who trust scientists working in natural science are very likely to also trust biologists. To test whether items are locally independent, researchers can examine the relationships between item responses after accounting for $\theta$ by performing residual correlation analysis [@chenLocalDependenceIndexes1997].

@@

To illustrate the application of GRM for analyzing measurement precision of an ordered polytomous scale, we provide an example of a GRM analysis of the RWA scale data [@altemeyerRightwingAuthoritarianism1981; @altemeyerAuthoritarians2006]. @altemeyerRightwingAuthoritarianism1981 defines RWA as a personality propensity to blindly abide by established authorities (i.e., authoritarian submission), to act aggressively toward individuals who are perceived to be punished by these established authorities (i.e., authoritarian aggression), and to uphold traditional values promoted by the authorities (i.e., conventionalism). Individuals with higher levels of RWA tend to view the world as a dangerous place [@duckittDualProcessMotivational2017], and thus, to ensure stability and safety, they are motivated to preserve social order by deferring to those whom they perceive as legal, social, or moral authorities; for instance, the government, religious institutions, or political or religious or military leaders [@saundersRightWingAuthoritarianismScale2017]. The RWA scale developed by @altemeyerRightwingAuthoritarianism1981 remains one of the most important measures in the field of social and political psychology [@saundersRightWingAuthoritarianismScale2017], and has helped researchers to conceptualize the role of RWA in shaping various social psychological outcomes, such as prejudice, political behavior, and various antisocial behaviors [@akramiRightWingAuthoritarianismSocial2006; @osbornePsychologicalCausesSocietal2023; @sibleyPersonalityPoliticalOrientation2012].

>>>>

---
title: "Getting Started with the Graded Response Model (GRM): An introduction and tutorial in R"
shorttitle: "The Graded Response Model Tutorial"
author:
  - name: Rizqy Amelia Zein
    corresponding: true
    orcid: 0000-0001-7840-0299
    email: amelia.zein@psikologi.unair.ac.id
    affiliations:
      - name: Ludwig-Maximilians-Universität
        department: Department of Psychology
        address: Leopoldstraße 13
        city: Munich
        region: Germany
        postal-code: "80802"
      - name: Universitas Airlangga
        department: Department of Psychology
        address: Jalan Airlangga 4-6
        city: Surabaya, Indonesia
        postal-code: "60286"
  - name: Hanif Akhtar
    orcid: 0000-0002-1388-7347
    affiliations:
      - name: ELTE Eötvös Loránd University
        department: Doctoral School of Psychology
        country: Hungary
      - name: Universitas Muhammadiyah Malang
        department: Faculty of Psychology
        city: Malang
        region: Jawa Timur, Indonesia
abstract: "This tutorial introduces the Graded Response Model (GRM), a tool for testing measurement precision within the Item Response Theory (IRT) paradigm, which is useful for informing researchers about the item and person properties of their measurement. The tutorial aims to guide applied researchers through a unidimensional GRM analysis in the R environment, using the psych, mirt, and ggmirt packages. GRM is specifically designed to examine the psychometric properties of psychological scales with polytomous (Likert-style) items. The tutorial illustrates the procedure using data from the Open Psychometrics Database on the Right-Wing Authoritarianism (RWA) scale, outlining the theoretical underpinnings of GRM and steps for data preparation, testing model assumptions, model fitting, plotting item parameters, and interpretation of results."
keywords:
  - graded response model
  - item response theory
  - mirt
  - ggmirt
  - psych
  - R
author-note:
  disclosures:
    conflict-of-interest: "We declare we have no competing interests."
    financial-support: "The first author receives a PhD scholarship from the German Academic Exchange Service (Deutscher Akademischer Austauschdienst, DAAD), award number 91803023."
lang: en
bibliography: references.bib
format:
  apaquarto-docx: default
editor: visual
execute: 
  warning: false
  error: false
  output: false
  echo: true
---

@@

library(flextable)
data <- read.csv("data/data.csv")
rename.items <- sprintf("%03d", 1:22)
colnames(data) <- rename.items
rm(rename.items)
```

To examine the quality of psychological measures, applied researchers implement statistical or formal mathematical models, which always require specific testable assumptions. Classical test theory (CTT) is one of the model-based measurement theories widely used to examine measurement precision, i.e., how accurate a measure is at capturing underlying psychological constructs and minimizing measurement error. While CTT is useful for informing researchers about measurement precision at the sample level, some want to look closely at specific individuals in their sample or items in their scale. In this sense, item response theory (IRT) fills this need by allowing researchers to inspect their measurement quality at the person and item level. IRT models are nonetheless known to be computationally intensive and less practical than CTT-based analysis, so a tutorial with didactic value can potentially help applied researchers to consider IRT analysis as a tool for their research.

The purpose of this tutorial is to provide a brief introduction to the graded response model (GRM), a family of IRT models specifically designed to assess the measurement precision of a polytomous (Likert-style) scale. In this paper, we aim to keep our tutorial concise, so we only briefly introduce the basics of IRT concepts. Interested readers are referred to available didactic texts on IRT, such as @embretsonItemResponseTheory2000 for the introductory level, and @bakerBasicsItemResponse2017 and @deayalaTheoryPracticeItem2022 for a more technical and comprehensive overview.

First, we briefly introduce readers to IRT and explain what GRM is and how the model is specified. In this part, we highlight how model and item parameters are specified and interpreted as well as visualizations derived from a GRM analysis. We also introduce three assumptions underlying a GRM and how to test these assumptions. We then provide an illustrative example of testing measurement precision using a GRM analysis, and in this part, we show how to practically implement a GRM analysis and interpret its results. We complement the illustrative example with code in the open-source programming language R [@rcoreteamLanguageEnvironmentStatistical2023] so that the readers can implement and reproduce the example presented in this paper. Readers with some experience in R and RStudio can easily follow this tutorial, but for those who are not yet familiar with R, we refer them to excellent, easy-to-follow materials developed by @navarroPsychologicalScience2018. Finally, we present some concluding remarks to encourage researchers to implement a GRM analysis as a part of their routine.

@@

To scrutinize their measurement quality, psychological researchers have relied on two theoretical frameworks underlying psychological testing, CTT and IRT, to decompose observed or raw scores into their *deterministic* (i.e., "true" score) and *random* (i.e., measurement error) components [@zumboValidityFoundationalIssues2006]. In practice, researchers implementing CTT principles apply structural equation modeling (SEM), exploratory factor analysis (EFA), or confirmatory factor analysis (CFA). Specifying the measurement model of an SEM model, or fitting EFA and CFA models, allows the covariance matrix of a given dataset to be decomposed into latent constructs ($\theta$) representing true scores or actual skill or trait levels, with measurement error (i.e., residuals or unique variances).

In this sense, CTT generally assumes that the relationship between observed score and true score is *linear*, as true score is essentially a linear transformation of observed score [@embretsonItemResponseTheory2000]. However, IRT models this relationship differently in a *probabilistic* rather than a linear fashion. More sharply, an IRT model assumes a probabilistic relationship between the observed score and the latent trait being measured ($\theta$). The probabilistic relationship in an IRT model accounts for the statistical properties of the items, such as item discrimination, difficulty, and, if necessary, pseudo-guessing parameter [@bakerBasicsItemResponse2017]. IRT methods were originally developed to evaluate traditional aptitude and achievement tests that measure knowledge and skill levels, but the application of IRT has been extended to personality, attitude, and other psychological scales of various formats.

The most critical difference between CTT and IRT is the way each framework conceptualizes measurement error. CTT uses standard deviation of the observed score (*s*) to calculate standard error of measurement. Because measurement error is based on the observed score, CTT assumes that measurement error is sample-dependent and constant for all individuals in the sample, regardless of their $\theta$ level. This shortcoming limits the utility of CTT, especially since psychologists often need to interpret individual scores, not just the test or the sample as a whole. In addition, because the standard error of measurement is also a function of reliability, it is assumed that scale reliability will always be the same across different levels of $\theta$. This is unrealistic because researchers often encounter situations in which a scale is "too hard" to endorse for a group of "low-performing" (i.e., low $\theta$) participants, providing little information beyond an indication that their $\theta$ level is much lower than what the scale can measure.

IRT offers a solution to this issue by enabling the calculation of standard error for each individual (*SE*$\theta$), and test reliability can be inferred from the average of *SE*$\theta$ across a sample of participants [@embretsonItemResponseTheory2000; @langSciencePracticeItem2021]. This approach allows for the estimation of reliability at varying levels of $\theta$ (i.e., test information function - TIF). By doing this, when a test is overly challenging for low-performing individuals or too easy for high-performing individuals, IRT analysis can help identify the levels of $\theta$ where the test is most reliable. A concrete example of this is a study scrutinizing the reliability of the Short Dark Tetrad (SD4) scale from an IRT perspective [@blotnerDarkTriadDead2022]. According to this study, the sadism subscale of the SD4 scale is the most reliable for measuring individuals with average to high levels ($\bar{x}$ \< $\theta$ \< 2.5 *SD*) of sadism but suboptimal for measuring those with low levels of sadism [@blotnerDarkTriadDead2022]. Therefore, IRT models offer a more informative approach, thus are useful for researchers who wish to closely examine the performance of their measures at the person level.

Further, IRT allows researchers to examine the performance of a specific item by specifying the relationship between item score and $\theta$ given one-, two-, or three-parameter. One-parameter logistic (1PL) or Rasch model accounts only varying item difficulty (*b*) while presuming equal item discrimination (*a*) across items. Practically, this model slightly overlaps with CTT in the sense that they assume that all items carry equal informative value thus the estimated $\theta$ of a 1PL/Rasch model is identical to the sum score of CTT [@langSciencePracticeItem2021; @stemlerRaschMeasurementItem2021]. In this sense, Rasch model assumes that the latent trait ($\theta$) should remain unaffected by specific items used in the test. For example, the difference in depression levels between two individuals should be always the same regardless of the scale used to measure their depression levels, be it Beck Depression Inventory (BDI) or Patient Health Questionnaire-9 (PHQ-9). While this principle reasonably enforces objectivity in measurement practices, it demands a strong theory delineating the construct and strict requirements of data-model fit, while both assumptions rarely hold in reality.

Furthermore, the two-parameter logistic (2PL) model fills this gap by allowing item ability to differentiate individuals with varying levels of $\theta$ (i.e., item discrimination parameter - *a*) to differ. Additionally, in some contexts, researchers may suspect that a part of the probabilistic relationship between observed score and $\theta$ is explained by guessing thus three-parameter logistic (3PL) model incorporates pseudo-guessing parameter (*c*).

Further, IRT paradigm has rapidly developed to include various models suited to specific contexts, such as handling ordinal responses [@murakiGeneralizedPartialCredit1992; @samejimaGradedResponseModel1997], categorical responses [@thissenNominalCategoriesItem2013], or assessing multidimensional traits simultaneously [@bockMarginalMaximumLikelihood1981; @chalmersMirtMultidimensionalItem2012]. It is important to note that 1PL, 2PL, and 3PL models are only applicable to binary or dichotomous data (e.g., true/false response), and the focus of this article is nonetheless to show the utility of IRT for fitting the ordered (Likert-style) responses. We briefly summarize the distinctions between 1PL, 2PL, 3PL, and GRM model in {apatb-table1}.

@@

```{r}
#| label: apatb-table1
#| apa-cap: Comparison Between Common IRT Models
#| apa-note: NULL
#| echo: false
```

| Model                       | Key Characteristics                                                                                                                                                                                                                                          | Data Type   | Response Options        |
|-----------------|--------------------|-----------------|-----------------|
| 1-PL Model (Rasch Model)    | 1\. Estimates only item difficulties (*b*). 2. Assumes that all items have equal discrimination parameters (*a*). 3. Item and person parameters are independent.                                                                                             | Dichotomous | Correct/Incorrect (0/1) |
| 2-PL Model                  | 1\. Estimates item difficulties (*b*) and item discrimination (*a*). 2. Less stringent than 1-PL model since it allows item discrimination parameters (*a*) to vary.                                                                                         | Dichotomous | Correct/Incorrect (0/1) |
| 3-PL Model                  | 1\. Estimates item difficulties (*b*), discrimination (*a*), and pseudo-guessing parameter (*c*). 2. Appropriate for modeling a test data with multiple responses (e.g., multiple-choice tests), and thus, guessing might influence participants' responses. | Dichotomous | Correct/Incorrect (0/1) |
| Graded Response Model (GRM) | 1\. Appropriate for modeling ordinal data with more than two response categories (i.e., Likert-style). 2. Estimates a discrimination parameter (*a*) and multiple threshold parameters (*b*) per item.                                                       | Polytomous  | Ordered Categories      |

@@

GRM is a family of IRT models specifically designed to analyze ordered polytomous (e.g., Likert-style) data [@samejimaGradedResponseModel1997; @samejimaGeneralGradedResponse2010; @samejimaGradedResponseModels2016]. GRM generally can be applied to a different types of test formats, such as scales measuring behavioral outcomes [@ariasHierarchyPsychometricProperties2016], personality and attitude scales [@rauthmannInvestigatingMACHIV2013], or patient-reported outcome measures [@normandGradedResponseModelbased2006]. Situational judgment tests (SJTs), which measure non-cognitive abilities, decision-making, and problem solving in context-specific situations [@corstjensSituationalJudgementTests2017], may also be suitable for GRM analysis when the response categories are ordered.

The fundamental idea of GRM is to extend the logic of simpler dichotomous IRT models to polytomous scales by applying a probabilistic function to each response category. In dichotomous two-parameter logistic (2PL) IRT models, the probability of answering a correct response is modeled as a function of $\theta$, item difficulty (*b*), and item discrimination (*a*). GRM extends this idea by modeling the probability of selecting certain response category or higher on an item (i.e., *step function*) given to item difficulty (*b*) and discrimination (*a*). For each response category (e.g., answering "agree" on a five-point Likert scale), GRM calculates the cumulative probability of a participant answering "agree" or above, given to their $\theta$. Therefore, the notion of item difficulty in dichotomous IRT models is extended to the step function (i.e., item threshold - *b*) to handle ordered polytomous data.

To illustrate the step function, consider a scale measuring sadistic personality (e.g., "*watching a fist-fight excites me*," etc.) with response categories ranging from 1 (*strongly disagree*) to 5 (*strongly agree*). Imagine these categories are hierarchically ordered from the lowest category (*strongly disagree*) to the highest (*strongly agree*), like a staircase. GRM aims to estimate the level of sadistic personality ($\theta$) required for "stepping" from one response category to another (e.g., *moderately agree* to *strongly agree*). Therefore, a GRM model of a five-point Likert scale calculates four item threshold parameters (*b*): the location of $\theta$ level where individuals are equally likely to respond 1 or 2 (*b*~1~), 2 or 3 (*b*~2~), 3 or 4 (*b*~3~), and 4 or 5 (*b*~4~). Each threshold is exactly the point at which a participant is equally likely to respond to either of the adjacent response categories. Note that the level of the latent trait ($\theta$) that a person has is typically centered around a mean of zero. The mean represents the average level of the trait across all participants in the sample. When we interpret item threshold parameters, such as b1, we are considering how far an individual's $\theta$ level deviates from this mean. For example, if *b*~1~ = -1.23 for the item "*watching a fist-fight excites me*", this means that participants with a sadism level 1.23 below the mean might change their response from *strongly disagree* to *moderately disagree*.

{apafg-fig1} shows an example of an item probability function (IPF) from an item on a scale with five response categories along with the staircase illustration. In general, IPF describes a relationship between $\theta$ (x-axis) and the probability of endorsing a response category (y-axis - *P*$\theta$). {apafg-fig1} consists of five category probability curves, each of which represents the probability of endorsing a response category given $\theta$. Item threshold is exactly the location of $\theta$ where two adjacent category probability curves cross each other. The "Strongly Disagree" and "Strongly Agree" curves are boundary categories in a GRM model, which results in their peaks being at the extreme ends of the $\theta$. These curves look different than the other because they represent the cumulative probability of selecting the most extreme responses, while intermediate categories peak in the middle and have more defined, bell-shaped curves. The probability curves of boundary categories simply mean that individuals at extreme $\theta$ levels are very likely to select extreme responses.

@@

```{r}
#| label: apafg-fig1
#| apa-cap: Staircase Illustration and Item Probability Functions of an Item in a GRM Model
#| apa-note: NULL
#| output: true
#| echo: false

@@

The second key assumption pertains to the unidimensionality of $\theta$. This implies that the item pool tested using a GRM analysis should represent only one latent trait ($\theta$). Although it is indeed possible to account for multiple $\theta$ in an IRT model (multidimensional IRT) we limit the scope of our tutorial to the unidimensional GRM. A unidimensional IRT model can still be robustly applied to multidimensional data if multiple latent traits are moderately intercorrelated or when there is a strong general factor (*g*) underlying the data [@reiseEvaluatingImpactMultidimensionality2015]. To test whether an item pool represents one $\theta$, researchers can apply a parallel analysis, a factor analysis, or a multidimensional scaling (MDS) [@deayalaAssessmentDimensionalityUse1991; @guoAssessingDimensionalityIRT2023], before specifying their GRM models.

Relatedly, a GRM model assumes that the item pool is locally independent, as the third assumption. Local independence suggests that participants' responses to one item do not affect their responses to another. To test whether items are locally independent, researchers can examine the relationships between item responses after accounting for $\theta$ by performing residual correlation analysis [@chenLocalDependenceIndexes1997]. In practice, it is quite difficult to fully satisfy the local dependency assumptions when researchers measure a construct with multi-item measures, so we should always expect some degree of local dependency to be present in the model. That said, some local dependency problems may be insignificant, and there are several strategies to avoid severe model misfit due to local dependency, which we will discuss later.

Note that evidence of unidimensionality does not always warrant local independence. As an eyeball example - consider a scale measuring trust in science that includes: "*I trust scientists working in natural science*" and "*Most biologists are trustworthy*." While both items may measure the same latent trait (i.e., trust in science), as indicated by moderate or large loading factors, those who trust scientists in the natural sciences are very likely to trust biologists as well. Therefore, in practice, it is likely that these items will still be moderately correlated even after taking into account their loading factors on $\theta$.

@@

To illustrate the application of GRM for analyzing measurement precision of an ordered polytomous scale, we provide an example of a GRM analysis of the RWA scale data [@altemeyerAuthoritarians2006]. @altemeyerAuthoritarians2006 defines RWA as a personality tendency to blindly obey established authorities (i.e., authoritarian submission), to act aggressively toward individuals who are perceived to be punished by these established authorities (i.e., authoritarian aggression), and to uphold traditional values promoted by authorities (i.e., conventionalism). Individuals with higher levels of RWA tend to view the world as a dangerous place, and thus, to ensure stability and security, they are motivated to maintain social order by deferring to those they perceive as legal, social, or moral authorities, such as the government, religious institutions, or political, religious, or military leaders [@saundersRightWingAuthoritarianismScale2017]. The RWA scale remains one of the most important measures in the field of social and political psychology and has helped researchers conceptualize the role of RWA in shaping various social psychological outcomes, such as prejudice, political behavior, and various antisocial behaviors.

In this paper, we use the most recent version of the RWA scale [@altemeyerAuthoritarians2006], which consists of 22 items in which participants are asked to indicate their agreement with the items on a nine-point scale ranging from "*strongly disagree*" (-4) to "*strongly agree*" (+4), with `0` as the midpoint. While the RWA scale consists of three sub-dimensions, i.e., Submissiveness, Aggression, and Conventionalism, the RWA scale is theoretically assumed to be unidimensional because @altemeyerAuthoritarians2006 shows that these sub-dimensions are strongly intercorrelated. While there is some evidence to the contrary [@duckittMultidimensionalityRightWingAuthoritarian2013] and despite the theoretical debates existing in the literature pertaining to the dimensionality of RWA, for the purposes of this article we will assume that the RWA construct is theoretically unidimensional.
